Hydras, Ordinals \& Co. \\ A library in Coq of entertaining formal mathematics

### 2021-10-04 doc/JFLA2022/paper.tex
macro for Coq-community

\usepackage{varioref}
\newtheorem{todo}{To do}
\usepackage{amsfonts, afterpage}

@@

\newcommand{\TODO}[2][]{[\textcolor{red}{TODO (#1):} \emph{#2}]}

@@

\begin{abstract}
  \emph{Hydra-battles} is a collaborative library on discrete math, written for the Coq proof assistant, and developed under the \emph{Coq-community} project. The proof scripts are
  accompanied with an electronic book, made with the help of the \emph{Alectryon} litterate proving tool.
  We present the evolution of the mathematical contents since
  former presentations at JFLA meetings.
  Then, we show how the structure of the project is determined   by two  requirements which must continuously satisfied:
   maintenance of the library with respect to several Coq-community projects it requires and evolutions of Coq, and 
  consistency of the book with respect to the library. 
\end{abstract}

@@

\begin{itemize}
\item formalizations of mathematics in proof assistants, such as Coq, continually grow in size and scope
\begin{itemize}
\item the Mathematical Components library and the four-color theorem and the odd order theorem in Coq
\item the Mathlib library for Lean
\end{itemize}
\item to continue to be useful as examples and reusable libraries, formalizations must be well \emph{documented} and \emph{maintained} to ensure they work with new proof assistant releases. %\TODO[Pierre]{add evolution of styles of proof and formalization (not limited to Coq litterature)? } 
\item Lean and Mathematical Components have accompanying books, which are written mainly to onboard new library users
\item book authors faced with several design choices: literate documents, using tools to keep the book and the proof assistant synchronized
\end{itemize}

\subsection{Vision}
The \emph{Hydra-battles} project, part of \emph{Coq-community}~\cite{CoqCommunity} aims to be an experimental platform for the collaborative development of commented libraries of formal proofs. \emph{Coq-community} is a community organization that we have founded in 2018 with two goals in mind: providing a solution for the long-term maintenance of interesting Coq packages, and working collaboratively on documentation projects. The \emph{Hydra-battles} project demonstrates that these two goals are not independent: interesting Coq packages can become the basis for new documentation.
%
This umbrella project now includes evolved versions of the former Cantor and Additions libraries (under the new names of Hydra-battles and Addition-chains), the Ackermann sub-library, extracted from Russel O'Connor's Goedel library~\cite{OConnor05, Goedel} and a bridge to the Gaia library (by José Grimm~\cite{Gaia,grimm:hal-00911710}).
%
By following this approach of commenting interesting Coq packages, we provide new documentation content, that contributes to the diversity of the thriving Coq documentation ecosystem.

We call on the Coq users in the JFLA community and beyond to come and join us in this effort, by bringing new interesting projects which are worth presenting to Coq learners, \emph{a.k.a.} Coq users, and guiding them in their exploration.
%
We also always have project ideas to extend further our explorations and anyone is welcome to join the team by sending small or larger contributions through pull requests.
%
The current state of the project is already the result of such evolutions after several of us contributed project solutions and new proposals to the initial version of the first author.

Futhermore, contrary to traditionally published books, the ``book'' that forms part of this project is intended to be forever evolving. As new Coq formalization patterns and proof techniques appear, the book can be adapted to demonstrate their use (in case they fit well with our applications).
%
By using modern maintenance techniques such as continuous integration and deployment, we can ensure that this documentation stays up to date with the latest Coq releases. With Alectryon~\cite{alectryonpaper, alectryongithub}, we ensure that code and documentation are always in sync.

@@

The contrast between the simplicity of the statements above and the complexity of their proofs convinced us that it is a good theme for a commented library~\cite{HydraBattles} of formal proofs written for the \textit{Coq} proof assistant~\cite{Coq}.

@@

 Prior stages of this project have already been presented at
 JFLA~\cite{PCiota, JFLA2018paper}.
We present recent evolutions of the library: new results, interaction within the Coq-community project~\cite{CoqCommunity}, and documentation generated with Alectryon~\cite{alectryonpaper, alectryongithub}.

@@

\section{Integration within Coq-community}

@@

\emph{Coq-community} is an informal organization run by volunteer Coq users with objectives to maintain interesting Coq projects in the long-term and help Coq users to collaborate on documentation, tooling, etc.

It was created in 2018, inspired by the \emph{Elm Community} organization~\cite{zimmermann:tel-02451322}.
%
Such ``Community Package Maintenance Organizations'' actually exist in many ecosystems as they avoid the very common problem of an important package becoming unmaintained as its author has moved on or disappeared~\cite{zimmermann2021grounded}.

In the case of Coq, this problem may be even more prevalent as many packages are created by PhD students, or by researchers for a specific paper, but their authors do not intend to maintain the package in the long-term.
%
The authors are however generally very open to having someone else who expresses interest in their work continue the maintenance.
%
\emph{Coq-community} makes that easy by providing a process for transferring or forking an unmaintained package, tooling for setting up good maintenance practices (such as continuous integration) and by making it possible for someone to take over a package without committing in the long-term (as maintainers who drop out can easily be replaced by some other volunteers).

Today, \emph{Coq-community} hosts over 50 projects maintained by over 30 maintainers.
%
The hosted projects come from a variety of origins.
%
Some had been maintained in the past by the Coq development team on behalf of the authors, but this meant that only minimal changes required to make the project build with new Coq versions were conducted.
%
Some were still maintained by their authors but they wanted help to keep them up-to-date.
%
Others were simply left unmaintained and were revived by their adoption in \emph{Coq-community}.

\subsection{Interactions with other Coq-community projects}

Given the objectives of \emph{Coq-community}, it makes sense to propose a transfer anytime we encounter an interesting project that is insufficiently maintained.
%
After the project is adopted, large changes and refactorings are explicitly allowed.
%
This means, for instance, that we can consolidate libraries together or split them into several packages if it makes sense.

In order to prove formally that the length of the considered
kind of battles is not given by a primitive recursive function, we used the formalization of primitive recursive functions, part
of Russell O'Connor's contribution on G\"{o}del first incompleteness theorem~\cite{OConnor05, Goedel}.
For this purpose, and above all by consideration of the scientific interest of this contribution, we chose to host and maintain this work within \textit{Coq-community}.

@@

\subsection{Documentation with Alectryon}

The hydra-battles book is written in LaTeX, but it makes very frequent references (about once per page, 274 snippets over 281 pages) to parts of the Coq development, showing definitions (Fig.~\ref{fig:Hprime}), lemmas (Fig.~\ref{fig:Ex42E0}), proof scripts (Fig.~\ref{fig:Ex42-schutte}), and computation results (Fig.~\ref{fig:Pow-evalPow17LetIn}). The order in which these references appear in the book is independent of the structure of our libraries. Thus, we chose not to include our textbook as comments in the libraries (solution \emph{à la} Coqdoc).
\TODO[Pierre]{is the preceding sentence OK?}

Originally, we copied snippets from Coq sources into LaTeX manually, and recorded and inserted the corresponding outputs manually as well.  This approach is common, but brittle: changes to Coq definitions or lemmas had to be reflected in the book's sources, and we found multiple instances where the book and the Coq development had diverged. \TODO[Clément]{see FIXME cite alectryon for a discussion of various approaches to documentation of Coq developments}

We solved this maintenance issue by moving to Alectryon, a tool to automatically record Coq proofs.  Instead of copy-pasting fragments into LaTeX, we now embed small LaTeX files automatically generated by Alectryon from our Coq development.  Our build system guarantees that these LaTeX snippets are always up-to-date and consistent with the code (and, by comparing these snippets across releases of Coq or versions of hydra-battles, we can easily spot unexpected changes).

The transition happened gradually, over a few weeks: for each snippet of Coq code that was in the book, we had to take the following steps:

\begin{enumerate}
\item Mark the snippet in the Coq sources (we use special comments \texttt{(* begin snippet \textit{name} *)} … \texttt{(* end snippet \textit{name} *)})
\item Configure output display, using special Alectryon annotations to configure what should be shown (only the inputs, inputs and outputs, some steps of the proof but not all, some proof states at key moments in a proof, etc.)
\item Replace the copy-pasted inputs and output in LaTeX with an \texttt{input} command.
\end{enumerate}

Importing snippets into a LaTeX document was not one of the original use cases of Alectryon, so we had to extend it: the original Alectryon did not have support for exporting to LaTeX, and it was towards documenting individual source files (a style where code and prose are interleaved within the same file and code is documented in the same order as it is compiled).  We extended Alectryon to support our needs by implementing a LaTeX backend, and by programming it to generate individual snippet files, one per \texttt{snippet} comment block.  The later part was straightforward: all it took was to build a custom Alectryon \textit{driver}, a small (about 100 lines) Python program that leverages most of the Alectryon toolchain but defines a custom frontend that understands our snippet annotations (and otherwise exposes the exact same command line as Alectryon).

@@

\emph{Hydra-battles} wants to be a bridge between scientific litterature (\emph{e.g.}~\cite{KP82, KS81, schutte} and proof assistant technology. For the mathematician, it may give a concrete view of the mathematical content, not only through full proofs, but also through significative computations: examples, functions associated with constructive proofs, etc. For the Coq user, it provides a consistent set of examples,
allowing to present and compare various formalization and proving techniques. It may be also used as a library, through opam packages.

It is also a medium sized library (more than 50K LOC), dependent on various tools and libraries of the Coq ecosystem. It may be also used to experiment new techniques of continuous maintenace of the code and its documentation.

@@

We plan to extend our libraries in two main directions.
The Gaia project~\cite{Gaia}, also maintained on \textit{Coq-community} contains a development by José Grimm~\cite{grimm:hal-00911710}. This library is dedicated  to the implementation in \textit{Coq (SSreflect)} of books from  Bourbaki's Elements of Mathematics. It contains data structures which are compatible with our ordinal notations. 
Our plan is to build a  bridge between  the combinatorial results of \textit{hydra-battles} and the set-theoretic content of \textit{Gaia}, and make possible the transfer of theorems between both libraries.
Another direction would be to write in \textit{Coq} a formal proof of the original statement of Theorem~\ref{kp:thm2}, using O'Connor's formalization of Peano Arithmetic~\cite{Goedel}.

>>>>

\usepackage{varioref}
\newtheorem{todo}{To do}
\usepackage{amsfonts, afterpage}
\usepackage{xspace}

@@

\newcommand{\TODO}[2][]{[\textcolor{red}{TODO (#1):} \emph{#2}]}
\newcommand{\coq}{Coq}
\newcommand{\community}{Coq-community\xspace}
\newcommand{\gaia}{Gaia}
\newcommand{\alectr}{Alectryon}
\newcommand{\equations}{Equations}

@@

\begin{abstract}
  \emph{Hydra-battles} is a collaborative library on discrete math, written for the {\coq}  proof assistant, and developed under the \community project. The proof scripts are
  accompanied with an electronic book, made with the help of the {\alectr} litterate proving tool.
  We present the evolution of the mathematical contents since
  former presentations at JFLA meetings.
  Then, we show how the structure of the project is determined   by two  requirements which must continuously satisfied:
   maintenance of the library with respect to several \community projects it requires and evolutions of {\coq}, and 
  consistency of the book with respect to the library. 
\end{abstract}

@@

\begin{itemize}
\item formalizations of mathematics in proof assistants, such as {\coq}, continually grow in size and scope
\begin{itemize}
\item the Mathematical Components library and the four-color theorem and the odd order theorem in {\coq}
\item the Mathlib library for Lean
\end{itemize}
\item to continue to be useful as examples and reusable libraries, formalizations must be well \emph{documented} and \emph{maintained} to ensure they work with new proof assistant releases. %\TODO[Pierre]{add evolution of styles of proof and formalization (not limited to Coq litterature)? } 
\item Lean and Mathematical Components have accompanying books, which are written mainly to onboard new library users
\item book authors faced with several design choices: literate documents, using tools to keep the book and the proof assistant synchronized
\end{itemize}

\subsection{Vision}
The \emph{Hydra-battles} project, part of \community~\cite{CoqCommunity} aims to be an experimental platform for the collaborative development of commented libraries of formal proofs. \community is a community organization that we have founded in 2018 with two goals in mind: providing a solution for the long-term maintenance of interesting {\coq} packages, and working collaboratively on documentation projects. The \emph{Hydra-battles} project demonstrates that these two goals are not independent: interesting {\coq} packages can become the basis for new documentation.
%
This umbrella project now includes evolved versions of the former Cantor and Additions libraries (under the new names of Hydra-battles and Addition-chains), the Ackermann sub-library, extracted from Russel O'Connor's Goedel library~\cite{OConnor05, Goedel} and a bridge to the Gaia library (by José Grimm~\cite{Gaia,grimm:hal-00911710}).
%
By following this approach of commenting interesting {\coq} packages, we provide new documentation content, that contributes to the diversity of the thriving {\coq} documentation ecosystem.

We call on the {\coq} users in the JFLA community and beyond to come and join us in this effort, by bringing new interesting projects which are worth presenting to {\coq} learners, \emph{a.k.a.} {\coq} users, and guiding them in their exploration.
%
We also always have project ideas to extend further our explorations and anyone is welcome to join the team by sending small or larger contributions through pull requests.
%
The current state of the project is already the result of such evolutions after several of us contributed project solutions and new proposals to the initial version of the first author.

Futhermore, contrary to traditionally published books, the ``book'' that forms part of this project is intended to be forever evolving. As new {\coq} formalization patterns and proof techniques appear, the book can be adapted to demonstrate their use (in case they fit well with our applications).
%
By using modern maintenance techniques such as continuous integration and deployment, we can ensure that this documentation stays up to date with the latest {\coq} releases. With {\alectr}~\cite{alectryonpaper, alectryongithub}, we ensure that code and documentation are always in sync.

@@

The contrast between the simplicity of the statements above and the complexity of their proofs convinced us that it is a good theme for a commented library~\cite{HydraBattles} of formal proofs written for the {\coq} proof assistant~\cite{Coq}.

@@

 Prior stages of this project have already been presented at
 JFLA~\cite{PCiota, JFLA2018paper}.
We present recent evolutions of the library: new results, interaction within the \community project~\cite{CoqCommunity}, and documentation generated with {\alectr}~\cite{alectryonpaper, alectryongithub}.

@@

\section{Integration within \community}

@@

\community is an informal organization run by volunteer {\coq} users with objectives to maintain interesting {\coq} projects in the long-term and help {\coq} users to collaborate on documentation, tooling, etc.

It was created in 2018, inspired by the {Elm Community} organization~\cite{zimmermann:tel-02451322}.
%
Such ``Community Package Maintenance Organizations'' actually exist in many ecosystems as they avoid the very common problem of an important package becoming unmaintained as its author has moved on or disappeared~\cite{zimmermann2021grounded}.

In the case of {\coq}, this problem may be even more prevalent as many packages are created by PhD students, or by researchers for a specific paper, but their authors do not intend to maintain the package in the long-term.
%
The authors are however generally very open to having someone else who expresses interest in their work continue the maintenance.
%
\community makes that easy by providing a process for transferring or forking an unmaintained package, tooling for setting up good maintenance practices (such as continuous integration) and by making it possible for someone to take over a package without committing in the long-term (as maintainers who drop out can easily be replaced by some other volunteers).

Today, \community hosts over 50 projects maintained by over 30 maintainers.
%
The hosted projects come from a variety of origins.
%
Some had been maintained in the past by the {\coq} development team on behalf of the authors, but this meant that only minimal changes required to make the project build with new {\coq} versions were conducted.
%
Some were still maintained by their authors but they wanted help to keep them up-to-date.
%
Others were simply left unmaintained and were revived by their adoption in \community.

\subsection{Interactions with other \community projects}

Given the objectives of \community, it makes sense to propose a transfer anytime we encounter an interesting project that is insufficiently maintained.
%
After the project is adopted, large changes and refactorings are explicitly allowed.
%
This means, for instance, that we can consolidate libraries together or split them into several packages if it makes sense.

In order to prove formally that the length of the considered
kind of battles is not given by a primitive recursive function, we used the formalization of primitive recursive functions, part
of Russell O'Connor's contribution on G\"{o}del first incompleteness theorem~\cite{OConnor05, Goedel}.
For this purpose, and above all by consideration of the scientific interest of this contribution, we chose to host and maintain this work within \community.

@@

\subsection{Documentation with {\alectr}}

The hydra-battles book is written in LaTeX, but it makes very frequent references (about once per page, 274 snippets over 281 pages) to parts of the {\coq} development, showing definitions (Fig.~\ref{fig:Hprime}), lemmas (Fig.~\ref{fig:Ex42E0}), proof scripts (Fig.~\ref{fig:Ex42-schutte}), and computation results (Fig.~\ref{fig:Pow-evalPow17LetIn}). The order in which these references appear in the book is independent of the structure of our libraries. Thus, we chose not to include our textbook as comments in the libraries (solution \emph{à la} Coqdoc).
\TODO[Pierre]{is the preceding sentence OK?}

Originally, we copied snippets from {\coq} sources into LaTeX manually, and recorded and inserted the corresponding outputs manually as well.  This approach is common, but brittle: changes to {\coq} definitions or lemmas had to be reflected in the book's sources, and we found multiple instances where the book and the {\coq} development had diverged. \TODO[Clément]{see FIXME cite alectryon for a discussion of various approaches to documentation of {\coq} developments}

We solved this maintenance issue by moving to {\alectr}, a tool to automatically record {\coq} proofs.  Instead of copy-pasting fragments into LaTeX, we now embed small LaTeX files automatically generated by {\alectr} from our {\coq} development.  Our build system guarantees that these LaTeX snippets are always up-to-date and consistent with the code (and, by comparing these snippets across releases of {\coq} or versions of hydra-battles, we can easily spot unexpected changes).

The transition happened gradually, over a few weeks: for each snippet of {\coq} code that was in the book, we had to take the following steps:

\begin{enumerate}
\item Mark the snippet in the {\coq} sources (we use special comments \texttt{(* begin snippet \textit{name} *)} … \texttt{(* end snippet \textit{name} *)})
\item Configure output display, using special {\alectr} annotations to configure what should be shown (only the inputs, inputs and outputs, some steps of the proof but not all, some proof states at key moments in a proof, etc.)
\item Replace the copy-pasted inputs and output in LaTeX with an \texttt{input} command.
\end{enumerate}

Importing snippets into a LaTeX document was not one of the original use cases of {\alectr}, so we had to extend it: the original {\alectr} did not have support for exporting to LaTeX, and it was towards documenting individual source files (a style where code and prose are interleaved within the same file and code is documented in the same order as it is compiled).  We extended {\alectr} to support our needs by implementing a LaTeX backend, and by programming it to generate individual snippet files, one per \texttt{snippet} comment block.  The later part was straightforward: all it took was to build a custom {\alectr} \textit{driver}, a small (about 100 lines) Python program that leverages most of the {\alectr} toolchain but defines a custom frontend that understands our snippet annotations (and otherwise exposes the exact same command line as {\alectr}).

@@

\emph{Hydra-battles} wants to be a bridge between scientific litterature (\emph{e.g.}~\cite{KP82, KS81, schutte} and proof assistant technology. For the mathematician, it may give a concrete view of the mathematical content, not only through full proofs, but also through significative computations: examples, functions associated with constructive proofs, etc. For the {\coq} user, it provides a consistent set of examples,
allowing to present and compare various formalization and proving techniques. It may be also used as a library, through opam packages.

It is also a medium sized library (more than 50K LOC), dependent on various tools and libraries of the {\coq} ecosystem. It may be also used to experiment new techniques of continuous maintenance of the code and its documentation.

@@

We plan to extend our libraries in two main directions.
The Gaia project~\cite{Gaia}, also maintained on \community, contains a development by José Grimm~\cite{grimm:hal-00911710}. This library is dedicated  to the implementation in {\coq} (SSreflect) of books from  Bourbaki's Elements of Mathematics. It contains data structures which are compatible with our ordinal notations. 
Our plan is to build a  bridge between  the combinatorial results of \textit{hydra-battles} and the set-theoretic content of \textit{Gaia}, and make possible the transfer of theorems between both libraries.
Another direction would be to write in {\coq} a formal proof of the original statement of Theorem~\ref{kp:thm2}, using O'Connor's formalization of Peano Arithmetic~\cite{Goedel}.
